In-silico molecular enrichment and clearance of the human intracranial space

### 2024-12-19 paper/supplementary_main.tex
Fix a bunch of labels and refs.

\subsection{Mixed-dimensional transport and flow equations}
\label{sec:more_maths}

@@

To complete~\eqref{eq:1d_pde}, we prescribe bifurcation conditions at
internal nodes of the perivascular networks $\Lambda_a, \Lambda_v$ and
boundary conditions at the end nodes. To this end, define the set of
internal nodes $\mathcal{B} \subset \mathcal{V}$ as the set of nodes
that are connected to two or more edges and the set of end nodes as
$\mathcal{N} \subset \mathcal{V}$, write $\hat{c}^i =
\hat{c}|_{\Lambda^i}$, and let $\mathcal{E}(y)$ denote the set of
(one, two or three) edges in $\{\Lambda^i\}$ sharing the node $y$. To
impose continuity of the concentrations, we set that the
concentrations when viewed from each edge must match at nodes:
\begin{equation}
  \hat{c}^i (y) = \hat{c}^j(y)
  \qquad \foralls \Lambda^i, \Lambda^j \in \mathcal{E}(y)
  \qquad \foralls y \in \mathcal{B} .
  \label{eq:conc_continuity}
\end{equation}
Moreover, to ensure \discuss{mass conservation}, we set that the flux
going in and out at nodes should add to zero, or more precisely, that
for each $y \in \mathcal{B}$:
\begin{equation}
  \sum_{\Lambda^i \in \mathcal{E}(y)}
  \left (\hat{D}^i A^i \partial_s \hat{c}^i (y) - A^i \hat u^i(y) \hat{c}^i(y) \right ) n^i(y)
  = 0
  \label{eq:flux_conservation}
\end{equation}
where the function $n^i : \mathcal{V} \rightarrow \{-1, 0, 1\}$
defines an orientation of each vertex relative to the
edges. Specifically, for an edge $\Lambda^i$ from $y_{\rm in}$ to $y_{\rm out}$
\begin{equation}
  n^i(y_{\rm in}) = 1, \quad
  n^i(y_{\rm{out}}) = -1, \quad
  n^i(y) = 0 \quad \foralls y \in 
  \mathcal{V} \backslash \{ y_{\rm in}, y_{\rm out} \} .  
  \label{eq:def:sign}
\end{equation}

@@

\subsubsection{Modelling CSF flow via the incompressible Stokes equations}
\label{sec:csf_flow}
The time-independent, incompressible Stokes equations model the flow
of an incompressible Newtonian fluid at low Reynolds numbers, and read
as follows: over the domain $\Omega_{\rm CSF} \subset \R^3$, find the
velocity vector field $\bm u : \Omega_{\rm CSF} \rightarrow \R^3$ and
the pressure field $p : \Omega_{\rm CSF} \rightarrow \R$ such that
\begin{subequations}
  \begin{alignat}{2}
    - \mu \Delta \bm u  + \nabla p &=  0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:momnetum_equation} \\ 
    \nabla \cdot \bm u &= 0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:divergence_equation} .
  \end{alignat}
  \label{eq:stokes}%
\end{subequations}%
In addition, we impose the following boundary conditions to model flow
induced by CSF production in the choroid plexus with $\Gamma_{\rm AM-U}$ as the main efflux site:
\begin{subequations}
  \begin{alignat}{2}
    \mu \nabla \bm u \cdot \bm{n} - p \bm n &= -R_0 ( \bm u \cdot \bm n ) \bm n
    && \text{ on } \Gamma_{\mathrm{AM-U}}, \\
    \bm u &= 0 && \text{ on } \Gamma_{\mathrm{AM-L}}\cup\Gamma_{\rm{pia}}\cup\Gamma_{\rm{SSAS}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{pia}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{SSAS}}, \\
    \bm u \cdot \bm n & = \frac{1}{ |\Gamma_{\rm LV}|}  u_{\rm in}, \quad \bm u \cdot \bm \tau = 0 \quad && \text{ on } \Gamma_{\rm{LV}} ,  
  \end{alignat}
  \label{eq:stokes_bcs}%
\end{subequations}%
where $\bm n$ denotes the unit outward normal to the boundary, $\mu$ is the (dynamic) CSF viscosity, $R_0 > 0$ is a resistance parameter for CSF efflux, and $u_{\rm in}$ is a given fluid influx, here across the lateral ventricle wall $\Gamma_{\rm LV}$. We consider alternative variations of these boundary conditions in connection with estimating dispersion effects induced by CSF pulsatility, see~\Cref{sec:supp:dispersion}.

@@

We solve the models presented in~\Cref{sec:more_maths} numerically
using finite element methods. The following subsections provide more
details, model-by-model.

@@

\subsection{Modelling dispersion in the CSF spaces}
\label{sec:supp:dispersion}

\mer{MER: Moved the below from Results.}
\mar{To examine cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm LV}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model (see \Cref{sec:csf_dispersion}).}

@@

Similar to the case of CSF flow driven by production introduced in \Cref{sec:csf_flow}, we employ a Stokes flow model to predict CSF velocities corresponding to the peak cardiac and respiratory blood inflow rates. In particular, we enforce a uniformly distributed inflow of $u^{\rm in}_{\rm AM}$ and $u^{\rm in}_{\rm LV}$ across the arachnoid mater ($\Gamma_{\rm AM}$) and lateral ventricular surfaces ($\Gamma_{\rm LV}$) respectively, mimicking the effect of reduced CSF volume due to blood inflow. We set $u^{\rm in}_{\rm AM} = 6\,$ml/s \cite{causemann2022human,baledent2014imaging} and $u^{\rm in}_{\rm LV}=0.31\,$ml/s \cite{vinje2019respiratory} in the cardiac case, and $u^{\rm{in}}_{\rm{AM}} = 1\,$ml/s \cite{gutierrez2022effect} and $u^{\rm in}_{\rm{LV}}=0.121\,$ml/s \cite{liu2024using} in the respiratory case, respectively.
We prescribe a no-slip boundary condition on the pial surface $\Gamma_{\rm Pia}$ and zero-traction on the spinal boundary $\Gamma_{\rm SSAS}$, making the spinal compartment the only outflow route of CSF. Thus, we compute the pulsatile flow fields by augmenting the Stokes flow model (eq. \ref{eq:momnetum_equation} and \ref{eq:divergence_equation}) with the following boundary conditions:
\begin{equation}\label{eq:cardiac_flow_bcs}
    \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
    \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
    \bm u = \frac{u^{\rm in}_{\rm LV} \cdot \bm n}{|\Gamma_{\rm LV}|} \text{ on } \Gamma_{\rm LV}; \quad
    \bm u = \frac{u^{\rm in}_{\rm AM} \cdot \bm n}{|\Gamma_{\rm skull}|} \text{ on } \Gamma_{\rm AM}.
\end{equation}
The obtained pressure fields $p_{\rm cardiac}$ and $p_{\rm resp}$ only account for viscous forces, albeit cardiac and respiratory induced CSF flow is known to be unsteady and substantially impacted by transient inertial forces. Assuming an oscillatory flow with either the heartbeat frequency $\omega_{\rm cardiac} = 2 \pi$ or respiratory frequency $\omega_{\rm resp} = 0.5 \pi$, a mean gap width of the CSF-filled spaces of $2h=3$\,mm and CSF density $\rho=10^3\,\text{kg/m}^3$, the squared Womersley number $\alpha_i^2 = \frac{h^2 \omega_{i} \rho}{\mu}$ becomes $\alpha_{\rm cardiac}^2 = 20.2$ and $\alpha_{\rm resp}^2=5.05$, which is unsteady. Based on theoretical considerations on the ratio of oscillatory flow to steady flow impedances in a tube, we account for inertial forces in the pulsatile flow by upscaling the computed viscous pressure field by a factor of $1 + \alpha^2 / 8$ (see \cite{van1998cardiovascular}, chapter. 4.3). Thus, the upscaled pressure fields become $p^{\rm up}_{i}= (1 + \alpha_{i}^2 / 8)p_i \text{ for } i = \{\mathrm{cardiac}, \mathrm{resp}\}$.

>>>>

\subsection{Mixed-dimensional transport and flow equations}
\label{sec:app:more_maths}

@@

To complete~\eqref{eq:1d_pde}, we prescribe bifurcation conditions at
internal nodes of the perivascular networks $\Lambda_a, \Lambda_v$ and
boundary conditions at the end nodes. To this end, define the set of
internal nodes $\mathcal{B} \subset \mathcal{V}$ as the set of nodes
that are connected to two or more edges and the set of end nodes as
$\mathcal{N} \subset \mathcal{V}$, write $\hat{c}^i =
\hat{c}|_{\Lambda^i}$, and let $\mathcal{E}(y)$ denote the set of
(one, two or three) edges in $\{\Lambda^i\}$ sharing the node $y$. To
impose continuity of the concentrations, we set that the
concentrations when viewed from each edge must match at nodes:
\begin{equation}
  \hat{c}^i (y) = \hat{c}^j(y)
  \qquad \foralls \Lambda^i, \Lambda^j \in \mathcal{E}(y)
  \qquad \foralls y \in \mathcal{B} .
  \label{eq:conc_continuity}
\end{equation}
Moreover, to ensure mass conservation, we set that the flux
going in and out at nodes should add to zero, or more precisely, that
for each $y \in \mathcal{B}$:
\begin{equation}
  \sum_{\Lambda^i \in \mathcal{E}(y)}
  \left (\hat{D}^i A^i \partial_s \hat{c}^i (y) - A^i \hat u^i(y) \hat{c}^i(y) \right ) n^i(y)
  = 0
  \label{eq:flux_conservation}
\end{equation}
where the function $n^i : \mathcal{V} \rightarrow \{-1, 0, 1\}$
defines an orientation of each vertex relative to the
edges. Specifically, for an edge $\Lambda^i$ from $y_{\rm in}$ to $y_{\rm out}$
\begin{equation}
  n^i(y_{\rm in}) = 1, \quad
  n^i(y_{\rm{out}}) = -1, \quad
  n^i(y) = 0 \quad \foralls y \in 
  \mathcal{V} \backslash \{ y_{\rm in}, y_{\rm out} \} .  
  \label{eq:def:sign}
\end{equation}

@@

\subsubsection{Modelling CSF flow via the incompressible Stokes equations}
\label{sec:csf_flow}
The time-independent, incompressible Stokes equations model the flow
of an incompressible Newtonian fluid at low Reynolds numbers, and read
as follows: over the domain $\Omega_{\rm CSF} \subset \R^3$, find the
velocity vector field $\bm u : \Omega_{\rm CSF} \rightarrow \R^3$ and
the pressure field $p : \Omega_{\rm CSF} \rightarrow \R$ such that
\begin{subequations}
  \begin{alignat}{2}
    - \mu \Delta \bm u  + \nabla p &=  0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:momentum_equation} \\ 
    \nabla \cdot \bm u &= 0 \quad && \text{ in } \Omega_{\rm CSF},
    \label{eq:divergence_equation} .
  \end{alignat}
  \label{eq:stokes}%
\end{subequations}%
In addition, we impose the following boundary conditions to model flow
induced by CSF production in the choroid plexus with $\Gamma_{\rm AM-U}$ as the main efflux site:
\begin{subequations}
  \begin{alignat}{2}
    \mu \nabla \bm u \cdot \bm{n} - p \bm n &= -R_0 ( \bm u \cdot \bm n ) \bm n
    && \text{ on } \Gamma_{\mathrm{AM-U}}, \\
    \bm u &= 0 && \text{ on } \Gamma_{\mathrm{AM-L}}\cup\Gamma_{\rm{pia}}\cup\Gamma_{\rm{SSAS}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{pia}}, \\
    %\bm u &= 0 && \text{ on } \Gamma_{\rm{SSAS}}, \\
    \bm u \cdot \bm n & = \frac{1}{ |\Gamma_{\rm LV}|}  u_{\rm in}, \quad \bm u \cdot \bm \tau = 0 \quad && \text{ on } \Gamma_{\rm{LV}} ,  
  \end{alignat}
  \label{eq:stokes_bcs}%
\end{subequations}%
where $\bm n$ denotes the unit outward normal to the boundary, $\mu$ is the (dynamic) CSF viscosity, $R_0 > 0$ is a resistance parameter for CSF efflux, and $u_{\rm in}$ is a given fluid influx, here across the lateral ventricle wall $\Gamma_{\rm LV}$. We consider alternative variations of these boundary conditions in connection with estimating dispersion effects induced by CSF pulsatility, see~\Cref{sec:app:dispersion}.

@@

We solve the models presented in~\Cref{sec:app:more_maths} numerically
using finite element methods. The following subsections provide more
details, model-by-model.

@@

\subsection{Modelling dispersion in the CSF spaces}
\label{sec:app:dispersion}

\mer{MER: Moved the below from Results.}
\mar{To examine cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm LV}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model.}

@@

Similar to the case of CSF flow driven by production introduced in \Cref{sec:csf_flow}, we employ a Stokes flow model to predict CSF velocities corresponding to the peak cardiac and respiratory blood inflow rates. In particular, we enforce a uniformly distributed inflow of $u^{\rm in}_{\rm AM}$ and $u^{\rm in}_{\rm LV}$ across the arachnoid mater ($\Gamma_{\rm AM}$) and lateral ventricular surfaces ($\Gamma_{\rm LV}$) respectively, mimicking the effect of reduced CSF volume due to blood inflow. We set $u^{\rm in}_{\rm AM} = 6\,$ml/s \cite{causemann2022human,baledent2014imaging} and $u^{\rm in}_{\rm LV}=0.31\,$ml/s \cite{vinje2019respiratory} in the cardiac case, and $u^{\rm{in}}_{\rm{AM}} = 1\,$ml/s \cite{gutierrez2022effect} and $u^{\rm in}_{\rm{LV}}=0.121\,$ml/s \cite{liu2024using} in the respiratory case, respectively.
We prescribe a no-slip boundary condition on the pial surface $\Gamma_{\rm Pia}$ and zero-traction on the spinal boundary $\Gamma_{\rm SSAS}$, making the spinal compartment the only outflow route of CSF. Thus, we compute the pulsatile flow fields by augmenting the Stokes flow model (\eqref{eq:momentum_equation} and \ref{eq:divergence_equation}) with the following boundary conditions:
\begin{equation}\label{eq:cardiac_flow_bcs}
    \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
    \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
    \bm u = \frac{u^{\rm in}_{\rm LV} \cdot \bm n}{|\Gamma_{\rm LV}|} \text{ on } \Gamma_{\rm LV}; \quad
    \bm u = \frac{u^{\rm in}_{\rm AM} \cdot \bm n}{|\Gamma_{\rm skull}|} \text{ on } \Gamma_{\rm AM}.
\end{equation}
The obtained pressure fields $p_{\rm cardiac}$ and $p_{\rm resp}$ only account for viscous forces, albeit cardiac and respiratory induced CSF flow is known to be unsteady and substantially impacted by transient inertial forces. Assuming an oscillatory flow with either the heartbeat frequency $\omega_{\rm cardiac} = 2 \pi$ or respiratory frequency $\omega_{\rm resp} = 0.5 \pi$, a mean gap width of the CSF-filled spaces of $2h=3$\,mm and CSF density $\rho=10^3\,\text{kg/m}^3$, the squared Womersley number $\alpha_i^2 = \frac{h^2 \omega_{i} \rho}{\mu}$ becomes $\alpha_{\rm cardiac}^2 = 20.2$ and $\alpha_{\rm resp}^2=5.05$, which is unsteady. Based on theoretical considerations on the ratio of oscillatory flow to steady flow impedances in a tube, we account for inertial forces in the pulsatile flow by upscaling the computed viscous pressure field by a factor of $1 + \alpha^2 / 8$ (see \cite{van1998cardiovascular}, chapter. 4.3). Thus, the upscaled pressure fields become $p^{\rm up}_{i}= (1 + \alpha_{i}^2 / 8)p_i \text{ for } i = \{\mathrm{cardiac}, \mathrm{resp}\}$.
